How environmental drivers of spatial synchrony interact

### 2023-05-31 Paper/Paper.Rnw
continued edits for referees

%nonlinearity
Our model is a linear population dynamical model, but most population dynamics are probably 
nonlinear, at least to some degree. We speculate on what may be the outcome if future work
explores interactions between two or more Moran effects using nonlinear population dynamical 
models. 
Some researchers have explored Moran effects of one environmental variable acting on nonlinear
population models and found generally that synchronous environmental fluctuations tend to promote
synchronous population fluctuations, in spite of nonlinearities, though often to a lesser 
degree than for the linear case \citep{Royama2005,Engen2005}. 
Our model is a linearization of a very general population dynamical model [SI section
S1.2 of \cite{Desharnais2018} and SI section S1 of \cite{Walter2017}], and it is well known
that analytic results for a linearized models, such as we have developed, also hold, approximately, for
the original nonlinear model, as long as the noise impinging the model is sufficiently weak. 
How weak is ``sufficiently weak'' is model dependent, but prior work examining that question
provided evidence that weak noise approximations using linearizations of commonly used 
ecological models are often actually quite accurate, even for fairly strong 
noise \cite{Nisbet1977}. 
Therefore we speculate that our main observations will hold even for nonlinear systems:
1) Moran effects can substantially interact, either synergistically or destructively;
2) interactions can depend on timescale; and 3) the nature of effects depend on relative,
timescale-specific lags of environmental influences on populations and between 
environmental variables. However, future work should nevertheless examine Moran interactions and 
nonlinear models, since much has been learned by exploring deviations of ecological models 
from predictions of theory based on linearization (e.g., \citealp{Reuman2006}).

%Conservation
Improved understanding of fundamental ecological phenomena often can be used to improve 
conservation efforts. But how, specifically, might the improved understanding of Moran effects that 
this study provides aid kelp conservation or restoration? Coauthor TWB is part of an effort
%***add detail on what effort, official names, maybe funders, etc. - whatever is appropriate to identify the project officially
to develop a stability model that informs the when, where and how of kelp restoration
following the 20XX crash of kelp populations in large areas along the west coast from X down to Y.
%***Tom to fill in the above missing info
Specifically, the stability model informs kelp restoration efforts by... 
%***Tom to fill in a very brief summary of how it works
It is well known that stability of regional populations is closely related to synchrony,
both generally \citep{Wilcox2017} and for kelp specifically \citep{Walter2021};
and the first step in the decision tree that TWB and his collaborators have developed 
involves computing the predicted stability of every subtidal reef across a large area.
%***Tom, we should cite your report, or other appropriate reference, somewhere around here.
Future iterations of TWB's and others' efforts to understand kelp stability and thereby 
support kelp restoration initiatives should incorporate consideration of interacting Moran
effects and their impacts on stability, and such consideration could likely be straightforwardly
integrated since existing methods already compute stability metrics as a key part of their process.

>>>>

%nonlinearity
Our model is a linear population dynamical model, but most population dynamics are probably 
nonlinear, at least to some degree. We speculate on what may be the outcome if future work
explores interactions between two or more Moran effects using nonlinear population dynamical 
models. 
Some researchers have explored Moran effects of one environmental variable acting on nonlinear
population models and found generally that synchronous environmental fluctuations tend to promote
synchronous population fluctuations, in spite of nonlinearities, though often to a lesser 
degree than for the linear case \citep{Royama2005,Engen2005}. 
Our model is a linearization of a very general population dynamical model [SI section
S1.2 of \cite{Desharnais2018} and SI section S1 of \cite{Walter2017}], and it is well known
that analytic results for linearized models, such as we have developed, also hold, approximately, for
the original nonlinear model, as long as the noise impinging the model is sufficiently weak. 
How weak is ``sufficiently weak'' is model dependent, but prior work examining that question
provided evidence that weak noise approximations using linearizations of commonly used 
ecological models are often actually quite accurate, even for fairly strong 
noise \cite{Nisbet1977}. 
Therefore we speculate that our main observations will hold even for nonlinear systems:
1) Moran effects can substantially interact, either synergistically or destructively;
2) interactions can depend on timescale; and 3) the nature of effects depend on relative,
timescale-specific lags of environmental influences on populations and between 
environmental variables. However, future work should nevertheless examine Moran interactions and 
nonlinear models, since much has been learned by exploring deviations of ecological models 
from predictions of theory based on linearization (e.g., \citealp{Reuman2006}).

%Conservation - new
A better understanding of fundamental ecological phenomena can be used to improve 
conservation efforts. But how, specifically, might improved understanding of Moran effects 
from this study aid kelp conservation or restoration? Scientists and managers are 
currently developing a stability model that informs when, where, and how to restore kelp 
forests following recent large-scale declines \citep{Bell2023}.
Specifically, these efforts inform kelp restoration by modeling the 
dynamics of kelp abundance (measured by diver surveys) as a function of environmental and 
bathymetric variables. Resulting maps of kelp stability are then compared to maps of recent  
canopy declines to identify sites where restoration 
activities are needed and would lead to stable future forests \citep{GiraldoOspinoInPrep}.
It is well known that stability of regional populations is influenced by synchrony, both 
generally \citep{Wilcox2017} and for kelp specifically \citep{Walter2021}; and the first step in 
the emerging kelp restoration decision tree of \cite{Bell2023} and \cite{GiraldoOspinoInPrep} 
involves predicting kelp stability across all 
suitable habitat in California. Future iterations of this process may consider 
interacting Moran effects to better predict stability and thereby better support kelp restoration.
